---
title: "Indexing Things in R"
author: "Mathew Kiang"
date: "2/1/2018"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

`R` has several ways of referring to parts of a data structure. They're all useful in different ways, but this process can be a bit confusing at first. Here is a set of references for indexing different data structures using different methods.

## Indexing a vector

You have some vector called `letter` which contains the (lowercase) letters of the alphabet.

```{r}
print(letters)
```

### By position

You can get the a single element by referring to its position (starting at 1). For example:

```{r}
print(letters[3])
```

You can refer multiple elements by passing a vector of their position. A continuous subset of elements is called a"slice" and the easiest way to slice something is to use the colon (`:`) operator. Here, we take the third to tenth letter, inclusive.

```{r}
print(letters[3:10])
```

The order of your indices matters. If I reverse the slice above, I get the same elements in reverse.

```{r}
print(letters[10:3])
```

You don't have to use slices. You can specify any subset of elements --- even the same one multiple times.

```{r}
print(letters[c(3, 3, 3, 3)])
```

```{r}
print(letters[c(13, 1, 20, 20)])
```

Note that above I used a single bracket (`[`). You can also use a double bracket ('[[`) when you are interested in only a single element. 

```{r}
print(letters[[3]])
```

### By Boolean vector

You can subset by providing a vector of logical values (`TRUE`/`FALSE`). If this indexing vector is shorter than the obect you are subsetting, it will get repeated. For example, suppse you want every third letter of the alphabet.

```{r}
print(letters[c(FALSE, FALSE, TRUE)])
```

This is useful when you don' tknow an element's position, but you are interested in some sort of rule. For example, you want all elements after "r":

```{r}
print(letters[letters > "r"])
```

Why does this work? Because `letters > "r"` returns a boolean vector:
```{r}
letters > "r"
```

### By removal

Negative indices indicate elements you want to *remove* from object. Note this is not typical of other programming languages (where negative means you are counting from the end). We can remove `c` by using a negative index.

```{r}
print(letters[-3])
```

### Getting the minimum/maximum/last element of a vector

Three common operations are extracting the maximum, minimum, or last object from an object. We can do that with `which.max()`, `which.min()`, and `length()`, respectively. Suppose you have a vector `nums` with some random set of integers.

```{r}
set.seed(1)
nums <- sample.int(20)
print(nums)
```

You can use the functions above to extract the element of interest. For example, the position of the maximum element is:

```{r}
which.max(nums)
```

We can use this to now index our original vector:

```{r}
nums[which.max(nums)]
```

`which()` is a general form:

```{r}
which(nums == 10)
```



<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Indexing lists in <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a>. Inspired by the Residence Inn <a href="http://t.co/YQ6axb2w7t">pic.twitter.com/YQ6axb2w7t</a></p>&mdash; Hadley Wickham (@hadleywickham) <a href="https://twitter.com/hadleywickham/status/643381054758363136?ref_src=twsrc%5Etfw">September 14, 2015</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>


```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
