---
title: "Hands-on intro to R"
subtitle: "(And maybe some modeling)"
author: "Mathew Kiang"
date: "1/24/2018"
output:
  xaringan::moon_reader:
    css: "slides_style.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Goals for today

1. Everybody has `R`, RStudio, and `deSolve` installed and can run code

--

2. Go through discrete time models of population and headaches

--

3. Intro to non-infectious ODE model of headaches

--

4. Modify an ODE model

---

class: center, middle, inverse
## Did anybody have trouble installing R?

---

class: center, middle, inverse
## Can you run this command with no errors?

```{r eval=FALSE}
library(deSolve)
```

---

class: center, middle, inverse
## Do you have the file from Canvas?
Download via Canvas

Also download from: https://goo.gl/2YDChe

---
# Tips

- In `RStudio` your code in the source panel (upper left) and use **CMD+Enter** or **Ctrl+Enter** to run it in the console
    - This allows you to build on old models without deleting your code

--

- Comment (`#`) your code as you build models (or as Caroline builds them)
    - Trust me, in a week you'll forget why you wrote something some way

--

- Drawing the model and/or writing the equations first helps
    - A lot

--

- You will be given boilerplate code
    - Always

--

- Use `?` to see the arguments for functions

--

- Use **ALT + -** to quickly make an assignment (<-)

--

- It'll be ok. We don't expect you to be `R` masters

---
# Discrete birth and deaths

- Here is `discrete_births_and_deaths.R` with comments removed:

```{r eval=FALSE}
population <- 1000
time_seq <- 1:500  
birth_rate <- 0.4
death_rate <- 0.4

dat <- NULL
for (t in time_seq) {
    new_population <- population + round(birth_rate*population) - 
        round(death_rate*population)

    dat <- rbind(dat, new_population)

    population <- new_population
}

plot(dat[, 1], type = "l")
```

- What's happening in this code? Line-by-line.
- What do we expect to see when we plot this code? Why?
- What do you need to do to make the line go up/down?

---

# Hands-on / Live coding

- Please form groups with 1-2 other students. Every group should have at least one person who went through the Swirl or SW tutorials.

---
class: center, middle, inverse
# That's it.
## Thanks