---
title: "Review: ODEs in R"
subtitle: "Like Wednesday — But Slower"
author: "Mathew Kiang"
date: "2016/1/27"
output:
  xaringan::moon_reader:
    css: "slides_style.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Goals for today

1. Review model code from Wednesday (line-by-line)

--

2. Visualize an SIR

--

3. Modify an SIR (make SIS, add births/deaths)

---
class: center, middle, inverse
## Can you run this command with no errors?

```{r eval=FALSE}
 library(deSolve)
```

--

## Do you have the file from Canvas?
Download via Canvas

Also download from: https://goo.gl/aSv6IO

---
# Tips

- In `RStudio` your code in the source panel (upper left) and use **CMD+Enter** or **Ctrl+Enter** to run it in the console
    - This allows you to build on old models without deleting your code

--

- Comment (`#`) your code as you build models (or as Caroline builds them)
    - Trust me, in a week you'll forget why you wrote something someway

--

- Drawing the model and/or writing the equations first helps
    - A lot

--

- You will be given boilerplate code
    - Always

--

- Use `?` to see the arguments for functions

--

- It'll be ok. We don't expect you to be `R` masters

---
class: center, middle, inverse
# SIR Code Review
## Let's go over each line

---
```{r eval=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
        
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I

        der <- c(dS, dI, dR)
        
         return(list(der))
     }) 
 }

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
Recall, this was the code for SIR models.

--

It is up on Canvas under Code.

---
```{r eval=FALSE}
 library(deSolve)

* parms <- c(beta = 0.333, k = 3 , r = 0.333)
* inits <- c(S = 499, I = 1, R = 0)
* dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
        
*       dS <- - (beta * k * S * I) / (S + I + R)
*       dI <- + (beta * k * S * I) / (S + I + R) - r * I
*       dR <- r * I

*       der <- c(dS, dI, dR)
        
         return(list(der))
     }) 
 }

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
You'll need to modify things that are highlighted

--

We will give you the rest (most of it).
---
```{r eval=FALSE}
* library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
        
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I

        der <- c(dS, dI, dR)
        
         return(list(der))
     }) 
 }

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
--
Just loads the `deSolve` library so you can use `ode()`
---
```{r eval=FALSE}
 library(deSolve)

* parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
        
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I

        der <- c(dS, dI, dR)
        
         return(list(der))
     }) 
 }

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
--

Make a vector of `parameter = value` pairs named `parms`

--

Every parameter you pass will require a value

--

You'll need to add/remove from `parms` as your model dictates

---
```{r eval=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
* inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
        
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I

        der <- c(dS, dI, dR)
        
         return(list(der))
     }) 
 }

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
--

Make a vector of `compartment = population` pairs named `inits`

--

Every compartment will need some initial value

--

Again, you'll need to add/remove from `inits` as your model changes

---
```{r eval=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
* dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
        
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I

        der <- c(dS, dI, dR)
        
         return(list(der))
     }) 
 }

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
--
Make a vector of time-steps named `dt` (equivalently, `dt <- 0:300`)

--

`seq(start, end, step)`

--

Want smaller time-steps? `seq(0, 300, .001)`

--

However, more time-steps (and finer time-steps) requires longer computational time

---
```{r eval=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

* SIR <- function(t, x, parms){
*     with(as.list(c(parms, x)), {
*       
*        dS <- - (beta * k * S * I) / (S + I + R)
*        dI <- + (beta * k * S * I) / (S + I + R) - r * I
*        dR <- r * I
* 
*       der <- c(dS, dI, dR)
*       
*         return(list(der))
*     }) 
* }

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```

Wrap up your model as a function

---
```{r eval=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

* SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
       
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I
 
       der <- c(dS, dI, dR)
       
         return(list(der))
     }) 
}

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
--

This `function` is named `SIR` and takes inputs `t`, `x`, `parms`

---
```{r eval=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
* dt <- seq(0, 300, 1)

* SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
       
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I
 
       der <- c(dS, dI, dR)
       
         return(list(der))
     }) 
}

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
`t` is the vector of time-steps

---
```{r eval=FALSE}
 library(deSolve)

* parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

* SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
       
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I
 
       der <- c(dS, dI, dR)
       
         return(list(der))
     }) 
}

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
`parms` is the vector of parameters

---
```{r eval=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

* SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
       
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I
 
       der <- c(dS, dI, dR)
       
         return(list(der))
     }) 
}

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
`x` is the current state of the model (required for `ode()` to work)

---
```{r eval=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

* SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
       
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I
 
       der <- c(dS, dI, dR)
       
         return(list(der))
     }) 
}

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
**NOTE:** Don't change the order. You can — but just don't. (Trust me.)

---
```{r eval=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
*     with(as.list(c(parms, x)), {
       
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I
 
       der <- c(dS, dI, dR)
       
         return(list(der))
     }) 
}

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
--

`c()` combines `parms` and `x` into one vector

--

`as.list()` converts that vector into a list

--

`with()` allows everything in the `{ }` to be referred to by shorthand

--

- Without `with()` you'd need to write `parms['k']`, `parms['beta']`, etc. every time

---
```{r eval=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
       
*        dS <- - (beta * k * S * I) / (S + I + R)
*        dI <- + (beta * k * S * I) / (S + I + R) - r * I
*        dR <- r * I
 
       der <- c(dS, dI, dR)
       
         return(list(der))
     }) 
}

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```
These lines define your compartments. 

--

Know how to modify these.

---
```{r eval=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
       
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I
 
*       der <- c(dS, dI, dR)
       
*        return(list(der))
     }) 
}

simulation <- as.data.frame(ode(y = inits, times = dt, 
                                func = SIR, parms = parms))
```

Collect compartments into `der` and `return` them as a `list`.

--

`ode()` needs the function to return **something** as a `list`

--

We just make a variable called `der` for convenience. We could do `return(list(c(dS, dI, dR)))`

---
```{r eval=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
       
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I
 
        der <- c(dS, dI, dR)
       
        return(list(der))
     }) 
}

* simulation <- as.data.frame(ode(y = inits, times = dt, 
*                                 func = SIR, parms = parms))
```
--

Run `ode()` on the model `SIR` with these `inits` and `parms`, for time `dt`

--

Then save that output as a `data.frame()` in a variable called `simulation`.

---
# What is in simulations?
```{r eval=TRUE, include=FALSE}
 library(deSolve)

 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
       
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I
 
        der <- c(dS, dI, dR)
       
        return(list(der))
     }) 
}

 simulation <- as.data.frame(ode(y = inits, times = dt, 
                                 func = SIR, parms = parms))
```

--

How many rows/columns?

--

What are they?

--
```{r}
head(simulation, 10)
```

--

(Could also run `simulation[1:10, ]`)

---
# Visualize it

```{r fig.width=10, fig.height=5}
matplot(x = simulation[, 1], y = simulation[, 2:4], type = "l")
```

--

`matplot()` just plots one column (`x`) against other columns (`y`) of a matrix.

---
# Visualize it (better)

```{r fig.width=10, fig.height=5}
matplot(x = simulation[, 1], y = simulation[, 2:4], type = "l", 
        lty = 1)
```

---
# Visualize it (better-er)

```{r fig.width=10, fig.height=5}
matplot(x = simulation[, 1], y = simulation[, 2:4], type = "l", 
        lty = 1, xlim = c(1, 40))
```

---
# Visualize it (better-er-er)

```{r fig.width=10, fig.height=5}
matplot(x = simulation[, 1], y = simulation[, 2:4], type = "l", 
        lty = 1, xlim = c(1, 40), 
        xlab = "Time", ylab = "People (count)", main = "SIR Model")
```

---
# Visualize it (best?)

```{r fig.width=10, fig.height=5}
matplot(x = simulation[, 1], y = simulation[, 2:4], type = "l", 
        lty = 1, xlim = c(1, 40), 
        xlab = "Time", ylab = "People (count)", main = "SIR Model")
legend(x = "right", legend = c('S', 'I', 'R'), col = 1:3, lty = 1)
```

---
# Visualize it

- See `?matplot` and `?legend` for more.

--

- Also can use `plot()` and `lines()` together.

--

For really pretty graphs, see `ggplot2`. 

- For example: https://mkiang.shinyapps.io/DiseaseDynamics/

---
class: center, middle, inverse
# Challenge Round
## Let's modify our models

---
# Make an SI model
```{r eval=FALSE, include=TRUE}
 parms <- c(beta = 0.333, k = 3 , r = 0.333)
 inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
       
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I
 
        der <- c(dS, dI, dR)
       
        return(list(der))
     }) 
}

 simulation <- as.data.frame(ode(y = inits, times = dt, 
                                 func = SIR, parms = parms))
```
What needs to be modified?

---
# Make an SI model
```{r eval=FALSE, include=TRUE}
* parms <- c(beta = 0.333, k = 3 , r = 0.333)
* inits <- c(S = 499, I = 1, R = 0)
 dt <- seq(0, 300, 1)

* SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
       
*        dS <- - (beta * k * S * I) / (S + I + R)
*        dI <- + (beta * k * S * I) / (S + I + R) - r * I
*        dR <- r * I
 
*        der <- c(dS, dI, dR)
       
        return(list(der))
     }) 
}

* simulation <- as.data.frame(ode(y = inits, times = dt, 
                                 func = SIR, parms = parms))
```
What needs to be modified?

---

```{r eval=FALSE, include=TRUE}
* parms_si <- c(beta = 0.333, k = 3)
* inits_si <- c(S = 499, I = 1)
 dt <- seq(0, 300, 1)

 SIR <- function(t, x, parms){
     with(as.list(c(parms, x)), {
       
        dS <- - (beta * k * S * I) / (S + I + R)
        dI <- + (beta * k * S * I) / (S + I + R) - r * I
        dR <- r * I
 
        der <- c(dS, dI, dR)
       
        return(list(der))
     }) 
}

 simulation <- as.data.frame(ode(y = inits, times = dt, 
                                 func = SIR, parms = parms))
```
Remove `R` from `parms` and `inits`.

Also renamed them so we don't overwrite old `parms` and `inits`.

---

```{r eval=FALSE, include=TRUE}
 parms_si <- c(beta = 0.333, k = 3)
 inits_si <- c(S = 499, I = 1)
 dt <- seq(0, 300, 1)

 SI <- function(t, x, parms){
     with(as.list(c(parms, x)), {

*       N <- S + I
        dS <- - (beta * k * S * I) / N
*       dI <- + (beta * k * S * I) / N
* 
    
*    der <- c(dS, dI)
       
        return(list(der))
     }) 
}

 simulation <- as.data.frame(ode(y = inits, times = dt, 
                                 func = SIR, parms = parms))
```
Remove `dR` from from our model.

Also define `N` in one place so we don't have to modify it multiple times.

---
```{r eval=FALSE, include=TRUE}
 parms_si <- c(beta = 0.333, k = 3)
 inits_si <- c(S = 499, I = 1)
 dt <- seq(0, 300, 1)

 SI <- function(t, x, parms){
     with(as.list(c(parms, x)), {

        N <- S + I
        dS <- - (beta * k * S * I) / N
        dI <- + (beta * k * S * I) / N

        der <- c(dS, dI)
       
        return(list(der))
     }) 
}

* simulation_si <- as.data.frame(ode(y = inits_si, times = dt, 
*                             func = SI, parms = parms_si))
```

Save simulations in a new object and change `ode()` call to our new `parms`, `inits`, and `func`.

---
```{r, include=FALSE}
parms_si <- c(beta = 0.333, k = 3)
inits_si <- c(S = 499, I = 1)
dt <- seq(0, 300, 1)

SI <- function(t, x, parms){
    with(as.list(c(parms, x)), {
        
        N <- S + I
        dS <- - (beta * k * S * I) / N
        dI <- + (beta * k * S * I) / N
        
        der <- c(dS, dI)
        
        return(list(der))
    }) 
}

simulation_si <- as.data.frame(ode(y = inits_si, times = dt, 
                                   func = SI, parms = parms_si))
```

```{r fig.width=10, fig.height=5}
matplot(x = simulation_si[, 1], y = simulation_si[, 2:3], type = "l", 
        lty = 1, xlim = c(1, 40), 
        xlab = "Time", ylab = "People (count)", 
        main = "SI Model")
legend(x = "right", legend = c('S', 'I'), 
       col = 1:2, lty = 1)
```

---
class: center, middle, inverse
# SIS
## With a neighbor, make a model that allows for returning from I to S.

---
## Example of SIS
```{r}
parms_sis <- c(beta = 0.333, k = 3, alpha = .3)
inits_sis <- c(S = 499, I = 1)
dt <- seq(0, 300, 1)

SIS <- function(t, x, parms){
    with(as.list(c(parms, x)), {
        
        N <- S + I
        dS <- - (beta * k * S * I) / N + (alpha * I)
        dI <- + (beta * k * S * I) / N - (alpha * I)
        
        der <- c(dS, dI)
        
        return(list(der))
    }) 
}

simulation_sis <- as.data.frame(ode(y = inits_sis, times = dt, 
                                   func = SIS, parms = parms_sis))
```

---

```{r fig.width=10, fig.height=5}
matplot(x = simulation_sis[, 1], y = simulation_sis[, 2:3], type = "l", 
        lty = 1, xlim = c(0, 40),
        xlab = "Time", ylab = "People (count)", 
        main = "SIS Model")
legend(x = "right", legend = c('S', 'I'), 
       col = 1:2, lty = 1)
```

---
class: center, middle, inverse
# SIR with births/deaths
## With a neighbor, make an SIR model where people can be born S and everybody dies
### Keep birth rate = death rate

---
## Example of SIR with births/deaths
```{r}
parms_bd <- c(beta = 0.333, k = 3 , r = 0.333, birth = .02, death = .02)
inits_bd <- c(S = 499, I = 1, R = 0)
dt <- seq(0, 300, 1)

SIR_bd <- function(t, x, parms){
    with(as.list(c(parms, x)), {
        
        N <- S + I + R
        dS <- - (beta * k * S * I) / N - (death * S) + (birth * N)
        dI <- + (beta * k * S * I) / N - r * I - (death * I)
        dR <- r * I - (death * R)
        
        der <- c(dS, dI, dR)
        
        return(list(der))
    }) 
}

simulation_bd <- as.data.frame(ode(y = inits_bd, times = dt, 
                                func = SIR_bd, parms = parms_bd))
```

---

```{r fig.width=10, fig.height=5}
matplot(x = simulation_bd[, 1], y = simulation_bd[, 2:4], type = "l", 
        lty = 1, xlim = c(1, 150), 
        xlab = "Time", ylab = "People (count)", 
        main = "SIR Model (with births/deaths)")
legend(x = "right", legend = c('S', 'I', 'R'), 
       col = 1:3, lty = 1)
```

---
# Conclusion

- As you can see, ODE models can get complex, very quickly
    - We could add births, return rates, seasonality, age structure, vaccination, vectors, changing human behavior, etc.

- Models will get harder than this, but you're beyond the steep learning curve at this point

- You'll need to know how to modify and build on them, but not necessarily the details of `R`
    - Point of using `R` is just to allow you to quickly see what happens to dynamics when you modify a model


---
class: center, middle, inverse
# That's it.
## Thanks