---
title: "Social Networks Worksheet"
author: "Mathew Kiang"
date: "3/6/2017"
output:
  xaringan::moon_reader:
    css: "slides_style.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, digits = 4)
```

class: center, middle
# Download the code.red[*]
## [https://git.io/vDSzh](https://git.io/vDSzh)

# Also download the data
## [https://git.io/vDSgT](https://git.io/vDSgT)

---
class: center, middle

# Question 1: Import a network, plot it, plot the degree distribution

---
## Import a network

```{r, include=FALSE}
## Load the libraries (install if necessary)
library(network)
library(tidyverse)
source('./code/network_practical.R')

## Load up data (make sure you have the right file path)
mat1 <- read_csv('./data/unnamed_contact_network.csv', trim_ws = TRUE)

## Convert the matrix into a network
net1 <- as.network(mat1, directed = FALSE)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
## Load the libraries (install if necessary)
library(network)
library(tidyverse)
source('./code/network_practical.R')

## Load up data (make sure you have the right file path)
mat1 <- read_csv('./data/unnamed_contact_network.csv', trim_ws = TRUE)

## Convert the matrix into a network
net1 <- as.network(mat1, directed = FALSE)
```

---
## Plot the network

```{r, fig.height=6, fig.width=9, message=FALSE, warning=FALSE}
## Take a look at the network
plot(net1, label = network.vertex.names(net1))
```

---
## Plot the degree distribution

```{r, fig.height=5, fig.width=10, message=FALSE, warning=FALSE}
deg <- NULL
for(i in 1:network.size(net1)) {
    deg <- rbind(deg, length(get.neighborhood(net1, i))) 
}
hist(deg)
```

---
class: center, middle

# Visualizing randomness
---
## 500 simulations plotted
```{r eval=FALSE}
## This just loops through n_sims times and makes that many lines/simulations
## NOTE: There's a difference between matplot and matlines
n_sims <- 500
for (i in 1:n_sims) {
    sim <- networkPractical(net1, .2, .2, runTime = 50)
    ## If it is the first simulation, we need to create the plot canvas
    if (i == 1) {
        matplot(sim$timeSeries, lty = 1, type = "l", xlim = c(0, 50), 
                ylim = c(0, network.size(net1)), col = alpha(1:3, .15))
    ## If it is not the first one, we just draw on top of the existing canvas
    } else {
        matlines(sim$timeSeries, lty = 1, type = "l", col = alpha(1:3, .15))
    }
}
```

Play with the parameters and see how this affects the plots.

---
## 500 simulations plotted
```{r, echo=FALSE, fig.height=6.5, fig.width=11}
## This just loops through n_sims times and makes that many lines/simulations
## NOTE: There's a difference between matplot and matlines
n_sims <- 500
for (i in 1:n_sims) {
    sim <- networkPractical(net1, .2, .2, runTime = 50, verbose = FALSE)
    ## If it is the first simulation, we need to create the plot canvas
    if (i == 1) {
        matplot(sim$timeSeries, lty = 1, type = "l", xlim = c(0, 50), 
                ylim = c(0, network.size(net1)), col = alpha(1:3, .15))
    ## If it is not the first one, we just draw on top of the existing canvas
    } else {
        matlines(sim$timeSeries, lty = 1, type = "l", col = alpha(1:3, .15))
    }
}
```

---
class: center, middle
# How does the starting infectious person change our model?

---

## Plot the network

```{r, fig.height=6, fig.width=9, message=FALSE, warning=FALSE}
vert_cols <- rep("black", network.size(net1))
vert_cols[c(7, 10)] <- c("green", "red")
plot(net1, label = network.vertex.names(net1), vertex.col = vert_cols)
```

Let's plot Node 7 (green) vs Node 10 (red). Note that Node 7 has only one edge, while Node 10 has 18. 

---

## Node 7 (green) vs Node 10 (red)
```{r, eval=FALSE}
n_sims <- 2000
for (i in 1:n_sims) {
    sim <- networkPractical(net1, .2, .2, runTime = 50, start_infected = 7)
    sim2 <- networkPractical(net1, .2, .2, runTime = 50, start_infected = 10)
    ## If it is the first simulation, we need to create the plot canvas
    if (i == 1) {
        matplot(sim$timeSeries[, 2], lty = 1, type = "l", xlim = c(0, 50), 
                ylim = c(0, 40), col = alpha(3, .03))
        matlines(sim2$timeSeries[, 2], lty = 1, type = "l", 
                 col = alpha(2, .03))
    ## If it is not the first one, we just draw on top of the existing canvas
    } else {
        matlines(sim$timeSeries[, 2], lty = 1, type = "l", col = alpha(3, .03))
        matlines(sim2$timeSeries[, 2], lty = 1, type = "l", 
                 col = alpha(2, .03))
    }
}
```

Modified the loop from before so that now we run two simulations -- one starting on Node 7 (green lines) and one starting on Node 10 (red lines). We plot only the infectious curve.

---
## Node 7 (green) vs Node 10 (red)
```{r, echo=FALSE, fig.height=6.5, fig.width=11}
n_sims <- 2000
for (i in 1:n_sims) {
    sim <- networkPractical(net1, .2, .2, runTime = 50, start_infected = 7, verbose = FALSE)
    sim2 <- networkPractical(net1, .2, .2, runTime = 50, start_infected = 10, verbose = FALSE)
    ## If it is the first simulation, we need to create the plot canvas
    if (i == 1) {
        matplot(sim$timeSeries[, 2], lty = 1, type = "l", xlim = c(0, 50), 
                ylim = c(0, 40), col = alpha(3, .03))
        matlines(sim2$timeSeries[, 2], lty = 1, type = "l", 
                 col = alpha(2, .03))
    ## If it is not the first one, we just draw on top of the existing canvas
    } else {
        matlines(sim$timeSeries[, 2], lty = 1, type = "l", col = alpha(3, .03))
        matlines(sim2$timeSeries[, 2], lty = 1, type = "l", 
                 col = alpha(2, .03))
    }
}
```

What does the variation between the two colors tell you? What does this imply for network structure and infectious diseases?
---

class: center, middle, inverse
# That's it.
## Thanks